Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2019-08-06 Response_to_Reviewers.tex

%Resubmit 15 Aug. 2019\\

\bigskip

We want to thank the two reviewers for their thorough evaluation and discussion of our paper, as well as the challenging questions they asked. We have revised the manuscript in a way that addresses most of the comments. In this response, we first want to address the two main questions that arose for both reviews. The first one (A) relates to the original mis-specification/ambiguities between the social and spatial components of the modelling process. The second one (B) relates to the contribution of the approach in terms of potential use for model makers, for domain experts and in operational contexts.

 \medskip
A. In terms of the ambiguity between the social and spatial component 
\begin{itemize}
\item A.1. We have precised and clarified the argument of the paper (mainly in introduction) by qualifying our contribution, which is to provide an operational framework for studying the influence of geometric structures of space on the results of a geosimulation model. We have thus distinguished two effects of geographical space in simulation modelling: the ones that are part of the model, reflecting for example the role played by distance on decreasing agents' perception or potential social interactions; and the ones which act as geometric structures "on which" the social elements simulated interact, i.e. geometric constraints of the simulation. The first type of effects is addressed by traditional modelling. Our original contribution is to tackle the second type. %\comment[CC]{done in 1.5 and 1.8}

\item A.2. We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e. groups of social agents whose behaviour is constrained by their position in geographical space. This term refers to empirical systems, which are the targets of the modelling process), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model. When we talk about the generator we actually used in the paper, we call it a "density grid generator" because the is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures. %\comment[CC]{done in 1.5}

 \item A.3. We have modified figure 1 to express the clarified approach visually. Naming the generator parameter "gamma" has helped reduced the confusion that the term 'meta-parameter' was inducing. We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper. %\comment[CC]{done}
\end{itemize}

@@

  \item B.3. As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper). We could also in another paper vary the size of grid and its cell shape (cf. MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example networks). \comment{il faut reprendre ces elements dans l'article}
\end{itemize}

\bigskip

\textbf{to do list:}

 \textbf{Sociospatial system} pour parler des systèmes empiriques. \comment[MLT]{Done}\\
 \textbf{interactions} pour parler de ce qui se passe dans le modele.\\
  \textbf{geometric structures} pour parler des outputs issus du générateur.\\
  + parler de "density grid generator" quand on développe le nôtre. \\
  caser" Jessop (2008) points in particular to the interdependence between social phenomena and the spatial environment where they take place. We model the social system as the interaction dynamics of agents and spatial units. The spatial generator produces the geometric support for these interactions." in intro 1.4 \\

@@

\textbf{First referee:}

\medskip

%The paper describes an approach that takes spatial sensitivity analysis explicitly and thoroughly into account. Based on the criticism that most agent-based-modelling and simulation approaches do neglect a careful recognition of the spatial impact on social phenomena, it introduces a “spatial generator” whose task is to generate different possible spatial configurations that are then used as a comparative measurement in socio-spatial simulations to prove the influence of spatial structures. To evaluate the strengths and weaknesses of the use of the spatial generator tool the author(s) applies two classical approaches – Schelling’s segregation model and the Sugarscape model – as benchmark models.

\begin{enumerate}
 % The methodology described is very convincing as it allows the utilisation of a tool that takes the spatial conditions – as causes or effects or intermediates – actually explicitly into account. With this approach, the paper convincingly outlines a pragmatic application by creating typical spatial compositions in urban contexts that is monocentric, polycentric or discontinuous cities.
 	\item \textit{From this methodological perspective of reviewing, it would be helpful to add some more information to approaches/techniques used within the tool of the spatial generator. For example, paragraph 2.8 mentions four dimensions of spatial structure but leave the reader alone with just the terms. Either describe these or omit it.}

   $\rightarrow$ To describe more the spatial generator we added a few sentences; and the generator was furthermore better situated among the many possible others in the discussion.
   %ways of making a spatial generator as Young2002 (FLN/MLT).
 	% item \comment[JR]{will add some description on urban form indicators etc.}

   \item \textit{This also applies to the three segregation indexes: what exactly do they measure?}

  $\rightarrow$ segregation indexes were better described in the section describing the models.
  %\comment[CC]{done in section "case study models"}

  \item \textit{The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added} \ldots
  %, for example, to delineate the terms “strength of attraction” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity? If so, is it then in space and/or time?).

  $\rightarrow$ \comment[JR]{will add more precision in the description of the generator}

  \item \textit{While the methodological part of the paper is convincing the theoretical and epistemological background is less so. In short, the paper would improve significantly, if most of the geographical and empirical associations would be left out.}
  %Even claimed by the author(s), the approach has not much to do with the empirical growth or development of cities, just because the approach reduces cities’ growth to the growth itself, omitting issues such as historical genesis, cultural differentiation, economic background political and legal circumstances, etc.

   $\rightarrow$ We have clarified the semantic terms used in sections 1.5 to 1.8 and in the figure 1.

   %\comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}\comment[MLT]{If we want to keep this part to please everyone (like our former reviewer), we could explicitly mention that we model the growth of a city (in terms of population density) which is the results of a bunch of dynamics ... and not the dynamics themselves?}
  %\comment[FLN]{ qui montre que la démarche méta est ici appliquée à un domaine empirique qu'il aide à explorer mais d'autres domaines peuvent s'emparer de la démarche méta.}

   % show disagreement with omitting empirical/geographical but demonstrate helps understand geographical question but the is generic and valuable for it.

  \item \textit{The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation. \ldots}
  %In so doing, the author(s) runs immediately into the problem of adequate definitions. In paragraph 1.2 the definition of geographical systems as “social agents interacting with one another, within a limited portion of space” is misleading. First, it is a tautology (geographical systems in space); secondly, geographical cannot be described with social agents, because geographical systems are composed of spatial units, their relations, and a spatially defined system border. Social systems, in turn, might be described with humans as elements, their relations, and again a socially defined system border. Moreover, with respect to sociological system theory (Luhmann 1993), space does not add any knowledge to the nature of social systems. This approach can be (and I would add: has to be) criticised, however, not by interming! ling fuzzily the social with the geographical, as happened with this definition. 
  %\item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ? which is far from geographical in the sense he puts it indeed}\comment[MLT]{use sociospatial system instead? The term is broadly used in the literature to describe the interdependence between social phenomena and their spatial environment. We could also quote \hyperlink{https://journals.sagepub.com/doi/abs/10.1068/d9107}{Jessop et al. 2008}}

 $\rightarrow$ cf. comment A

  %\item reprendre tout le papier pour éliminer les termes indéfinis "space", "pattern" etc. + en conclu mentionner que ça s'applique aux automates cellulaires (sans agents) et aux modèles écologiques (agents-arbres).
  %\item in conclusion, we need to clarify the answer to the question "do geometric structure influence sociospatial systems dynamics" by saying it depends on the model. ≠ black\&white and we can demonstrate it. The nature of social systems is a given (in the model mu, which is not the contribution of this paper) and we explore the consequence of geometric structures on the spatial expression of social systems.

  \item \textit{Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.} \ldots
  % Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account. These interactions can then be associated with the material and relational dimension of geographical spaces. The question remains: what is the role of geographical systems? In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.

  $\rightarrow$ see comment A. cite Banos, 2012

  \textit{The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following. What is described here as "spatial system shape" is the geometry of spaces but not the geography of spaces.}
  % -> yes indeed

  $\rightarrow$ see comment A.

  \item \textit{Even more critically is the arbitrary conflation of social and spatial units as happened in paragraph 2.1, as geographical entities are described with “people, housing, networks, etc.” One is inevitably asking oneself: do geographical entities comprise of simply everything? Obviously not, but the paper seems to argue this way.}

  $\rightarrow$ All these terms were clarified (see comment A).

  \item \textit{Due to this, attempts to explain the growth of existing cities should be avoided.}

  $\rightarrow$ We agree with this comment and recall that urban growth processes are not the main focus of the paper, they are here an instantiation of a given $\gamma$ generator, which could have been replaced by others such as network models, procedural modeling, etc. The position and discussion of the paper was amended to precise that (see comment A).

  %it's not the point of the paper, only the metaphore of the spatial generator. + supprimer la ref à "urban growth" dans le texte et dire d'inspiration urban science mais dont seules les grilles finales nous intéressent. On aurait pu utiliser des generator issus de graphes, de percolation etc.
  %The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
  %\item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.

@@

%This manuscript presents a methodology to explore the effects of spatial parameters in ABM. They basically explore what they called “generator parameters”, to distinguish them from other statistical ones, which include density spatial distribution and patterns. To do so, they compared phase diagrams and calculated the mean of final values. 
%As case-studies, they used the actual distributions of three European cities as a means to constraint the density generation in two archetypical model (Schelling's segregation model and Epstein and Axtell's Sugarscape).

\medskip

\begin{enumerate}

  \item \textit{While the different outcomes of the two implementation exercises are not the main point, I would suggest authors to provide some more elaborated explanations. The two models have different behavioural assumptions and represent a different 'interaction' context. How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes?}

  $\rightarrow$ see above comment A, discussion has been amended and details added.
  \medskip

  \item \textit{In 4.1, the authors suggest potential developments and point to certain limitations. While limitations are well-discussed, readers who are not specialist of spatial models could benefit from a more comprehensive discussion on potential applications and extensions.}

  $\rightarrow$ see comment B

  \textit{For instance, it would be useful to have suggestions on how to use this methodology to run empirical models. Suppose to have a nice dataset and a case of a urban context to be modelled. Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?}

  $\rightarrow$ cf. comment B.3
  % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}

@@

  $\rightarrow$ cf. comment B.3
  % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}

  \item it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig. 1). We thank reviewer 2 for highlighting this limitation.

  \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}

  $\rightarrow$ cf. discussion of egohoods in conclusion (comment B)

@@

  $\rightarrow$ cf. discussion of egohoods in conclusion (comment B)

  \medskip

  %Spatial econometrics based on empirical dataset is developing fast. It would be useful to understand how these models can enrich ABM and vice-versa. 
   \item this is very interesting to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010. We need to add a paragraph about this discussion in conclusion and how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood'). cf. comment B

  \item\textit{Furthermore, even among limitations, perhaps authors could discuss data limitations and the fact that social processes (networks, mobility) cannot conflate with spatial dynamics. Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones!}

  $\rightarrow$ we have clarified this issue in introduction but referring to different terms and figure 1 which differentiates between the mu model and gama parameters. (see comment A)

  \item\textit{In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.} \ldots
  %I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints. And perhaps, authors could suggest future line of research that could show-case their argument.

  $\rightarrow$ cf. new discussion/conclusion and comment B

%I appreciate that models are fully available to readers. 
%\item We do too!

### 2019-08-06 paper.tex
.

@@

Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.
% Generators of abstract urban systems at a macroscopic scale could also be useful tools to better understand urban growth models as sketched by \cite{raimbault2018unveiling}.}

@@

>>>>

\subsubsection{Sensitivity of indicators to the geometry}

While we investigated the sensitivity to geometry of models together with their indicators, a potential research direction is to study the dependance of indicators themselves to the spatial configuration. % TODO finish

@@

Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2019urban}. The particular density grid generator we used here is an example among possible others, and our methodological contribution is generic as it does not depend on which generator is chosen.

@@

This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}. % TODO finish
